---
output:
    pdf_document:
      citation_package: natbib
      df_print: tibble
      fig_caption: yes
      keep_tex: no
      template: "../inst/extdata/markdown/latex-ms.tex"
title: 'SITS: Data Analysis and Machine Learning for Data Cubes using Satellite Image Time Series'
author:
- affiliation: National Institute for Space Research (INPE), Brazil
  name: Rolf Simoes
- affiliation: National Institute for Space Research (INPE), Brazil
  name: Gilberto Camara
- affiliation: National Institute for Space Research (INPE), Brazil
  name: Pedro R. Andrade
- affiliation: Institute for Applied Economics Research (IPEA), Brazil
  name: Alexandre Carvalho
- affiliation: National Institute for Space Research (INPE), Brazil
  name: Lorena Santos
- affiliation: National Institute for Space Research (INPE), Brazil
  name: Karine Ferreira
- affiliation: University of Vienna, Austria
  name: Victor Maus
- affiliation: National Institute for Space Research (INPE), Brazil
  name: Gilberto Queiroz
date: "`r format(Sys.time(), '%B %d, %Y')`"
endnote: false
fontfamily: mathdesign
fontfamilyoptions: adobe-utopia
fontsize: 11pt
graphics: true
mathtools: true
bibliography: ../inst/extdata/markdown/references-sits.bib
abstract: Using time series derived from big Earth Observation data sets is one of the leading research trends in Land Use Science and Remote Sensing. One of the more promising uses of satellite time series is its application for classification of land use and land cover, since our growing demand for natural resources has caused major environmental impacts. Here, we present an open source *R* package for satellit eimage time series analysis called `sits`. Package `sits` provides support on how to use statistical learning techniques with image time series obtained from data cubes. These methods include linear and quadratic discrimination analysis, support vector machines, random forests, deep learning and neural networks.
vignette: |
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteIndexEntry{SITS: Data Analysis and Machine Learning using Satellite Image Time Series}
---
```{r, include = FALSE}
library(sits)
library(tibble)
library(dtwclust)
```

# Introduction 

Earth observation satellites provide a regular and consistent set of information about the land and oceans of the planet. Recently, most space agencies have adopted open data policies, making unprecedented amounts of satellite data available for research and operational use. This data deluge has brought about a major challenge: *How to design and build technologies that allow the Earth observation community to analyse big data sets?*

The approach taken in the current work is to develop data analysis methods that work with satellite image time series, obtained by taking calibrated and comparable measures of the same location in Earth at different times. These measures can be obtained by a single sensor (e.g., MODIS) or by combining different sensors (e.g., Landsat 8 and Sentinel-2). If obtained by frequent revisits, the temporal resolution of these data sets can capture important land use changes. 

Time series of remote sensing data show that land cover can occur not only in a progressive and gradual way, but they may also show discontinuities with abrupt changes [@Lambin2003]. Analyses of multiyear time series of land surface attributes, their fine-scale spatial pattern, and their seasonal evolution leads to a broader view of land-cover change. Satellite image time series have already been used in applications such as mapping for detecting forest disturbance [@Kennedy2010], ecology dynamics [@Pasquarella2016], agricultural intensification [@Galford2008], and its impacts on deforestation [@Arvor2012]. Algorithms for processing image time series include BFAST for detecting breaks [@Verbesselt2010], TIMESAT for modelling and measuring phenological attributes [@Jonsson2004] and methods based on Dynamic Time Warping (DTW) for land use and land cover classification [@Petitjean2012][@Maus2016]. 

In this work, we present SITS, an open source R package for satellite image time series analysis. It provides support on how to use machine learning techniques with image time series. These methods include linear and quadratic discrimination analysis, support vector machines, random forests, and neural networks. One important contribution of the SITS package is to support the complete cycle of data analysis for time series classification, including data acquisition, visualisation, filtering, clustering, classification, validation and post-classification adjustments. 

Most studies using satellite image time series for land cover classification use a \emph{space-first, time-later} approach. For multiyear studies, researchers first derive best-fit yearly composites and then classify each composite image. For a review of these methods for land use and land cover classification using time series, see [@Gomez2016]. As an alternative to \emph{Space-first, time-later} methods, the SITS package provides support for classification of time series, preserving the full temporal resolution of the input data, using a \emph{time-first, space-later} approach. SITS uses all data in the image time series to create larger dimensional spaces for machine learning. The idea is to have as many temporal attributes as possible, increasing the dimension of the classification space. Each temporal instance of a time series is taken as an independent dimension in the feature space of the classifier. To the authors' best knowledge, the classification techniques for image time series included in the package are not previoulsy available in other R or python packages. Furthermore, the package includes methods for filtering, clustering and post-processing that also have not been published in the literature. 

# Image data cubes as the basis for big Earth observation data analysis

In broad terms, the cloud computing model is one where large satellite-generated data sets are archived on cloud services, which also provide computing facilities to process them. By using cloud services, users can share big Earth observation databases and minimize the amount of data download. Investment in infrastructure is minimised and sharing of data and software increases. However, data available in the cloud is best organised for analysis by creating data cubes. 

Generalising @Appel2019, we consider that a data cube is a four-dimensional structure with dimensions x (longitude or easting), y (latitude or northing), time, and bands. Its spatial dimensions refer to a single spatial reference system (SRS). Cells of a data cube have a constant spatial size (with regard to the cube’s SRS). The temporal dimension is specified by a set of intervals. For every combination of dimensions, a cell has a single value. Data cubes are particularly amenable for machine learning techniques; their data cane be transformed into arrays in memory, which can be fed to training and classification algorithms. Given the widespread availability of large data sets of Earth observation data, there is a growing interest in organising large sets of data into "data cubes".

As explained below, a data cube is the data type used in `sits` to handle dense raster data. Many of the operations involve creating, transforming and analysing data cubes.

## Using Web Data Services to Access Image Data Cubes

One of the distinguishing features of SITS is that it has been designed to work with big satellite image data sets which reside on the cloud and with data cubes. Many *R* packages that work with remote sensing images require data to be accessible in a local computer. However, with the coming of age of big Earth observation data, it is not always practical to transfer large data sets. Users have to rely on web services to provide access to these data sets. In this context, SITS is based on access to data cubes using web services.

A web service is software system designed to support remote access to image collections through APIs. These are machine-to-machine protocols that allow access to image collections and to generate *data cubes*. SITS uses two kinds of services: those that provide time series data and those that provide access to data cubes. Currently, there are no established standards for these two services. For this reason, `sits` uses two new protocols: WTSS ("Web Time Series Serice") and EOCubes ("Earth observation Data Cubes Services") that we have develop.  WTSS is a light-weight service for retrieval of time series data for selected locations and periods [@Vinhas2016]. The WTSS R client is available in the CRAN archive using the `wtss` package. The EOCUBES service provides access to big data collections, which are organised by the `EOCubes` package. the package also allows users to work on local files, if desired.

To find out the services and respective data cubes available, one should use the `sits_services()` function. 
```{r}
sits_services()
```

These services are set on the SITS configuration file, which is described later in this document. For each services, the above function lists the names of the data cubes available and additional information. 

## Defining a data cube using the WTSS service

To define a data cube for the WTSS, in principle the following parameters should be provided: (a) the name of the service; (b) the URL; (c) the satellite and sensor associated to the cube, and (d) name of the data cube in the remote service. If the URL, satellite and sensor are not provided, the package will search for default information in the SITS configuration file. 

```{r}
# define the data cube "MOD13Q1" using the WTSS service
# In this case, the WTSS service is run by a server in INPE Brazil
wtss_cube <- sits_cube(service = "WTSS", 
                       name    = "MOD13Q1")

# get information on the data cube 
wtss_cube %>% dplyr::select(service, URL, satellite, sensor)
# spatial dimensions of the data cube
wtss_cube %>% dplyr::select(xmin, xmax, ymin, ymax)
# temporal dimension of the data cube
timeline <- sits_timeline(wtss_cube)
message(paste0("start date = ", timeline[1], " end date = ", timeline[length(timeline)], " steps = ", length(timeline)))
# bands of the data cube
sits_bands(wtss_cube)
```

## Defining a data cube using the EOCUBES service

In the case of EOCUBES service, additional obtional parameters are: (a) a string with tile names; (b) a geometry to restrict the spatial extent; (c) start and end dates to restrict the temporal extent. If desired, the cube based on the EOCUBES service can be specified only by the service and cube names, provided that the correct additional information is provided in the configuration file.

```{r}
# create a coverage from EOCUBES service from the collection "MOD13Q1/006"
modis_cube <- sits_cube(service = "EOCUBES",
                        name    = "MOD13Q1/006")

# get information on the data cube 
modis_cube %>% dplyr::select(service, URL, satellite, sensor)
# get information on the cube
modis_cube %>% dplyr::select(xmin, xmax, ymin, ymax, timeline)
```

# Defining a data cube using files organised as raster bricks

The SITS package enables uses to create data cube based on files. In this case, these files should be organized as `raster bricks`. A RasterBrick is a multi-layer raster object used by the \emph{R} `raster` package.  Each brick is a multi-layer file, containing different time instances of one spectral band. To allow users to create data cubes based on files, SITS needs to know what is the timeline of the data sets and what are the names of the files that contain the RasterBricks.  The example below shows two bricks which are available at the AWS service, each containing 392 time instances of the "ndvi" and "evi" bands for the years 2000 to 2016. The timeline is available as part of the SITS package.

In the example, neither the satellite nor the sensor are provided; this information is deduced by SITS considering the cartographical projection associated to the image. If the projection is "sinusoidal", SITS will deduce these are MODIS files; data on "utm" projection is associated to the LANDSAT-8 satellite. 

```{r}
# create a coverage from files available in AWSx
ndvi_file <- paste0("/vsicurl/https://s3-sa-east-1.amazonaws.com/landsat-modis/Sinop_ndvi.tif")
evi_file <- paste0("/vsicurl/https://s3-sa-east-1.amazonaws.com/landsat-modis/Sinop_evi.tif")

# define the timeline
data(timeline_modis_392)

# create a raster metadata file based on the information about the files
raster_cube <- sits_cube(name = "Sinop", timeline = timeline_modis_392, bands = c("ndvi", "evi"),
                         files = c(ndvi_file, evi_file))

# get information on the data cube 
raster_cube %>% dplyr::select(service, URL, satellite, sensor)
# get information on the coverage
raster_cube %>% dplyr::select(xmin, xmax, ymin, ymax)
```

Once a data cube has been created, it can be used for extracting time series and for data analysis and classification, using the function `sits_get_data()`, descibed below. 


# Data structures for satellite image time series

The `sits` package requires a set of time series data, describing properties in spatio-temporal locations of interest. For land use classification, this set consists of samples provided by experts that take *in-situ* field observations or recognize land classes using high resolution images. The package can also be used for any type of classification, provided that the timeline and bands of the time series (used for training) match that of the data cubes. 

For handling time series, the package uses a `sits tibble` to organize time series data with associated spatial information. A `tibble` is a generalization of a `data.frame`, the usual way in *R* to organise data in tables. Tibbles are part of the `tidyverse`, a collection of R packages designed to work together in data manipulation [@Wickham2017]. As a example of how the `sits` tibble works, the following code shows the first three lines of a tibble containing $2,115$ labelled samples of land cover in Mato Grosso state of Brazil. It is the most important agricultural frontier of Brazil and it is the largest producer of soybeans, corn, and cotton. The samples contain time series extracted from the MODIS MOD13Q1 product from 2000 to 2016, provided every $16$ days at $250$-meter spatial resolution in the Sinusoidal projection. Based on ground surveys and high resolution imagery, it includes $425$ samples of nine classes: "Forest", "Cerrado", "Pasture", "Soybean-fallow", "Fallow-Cotton", "Soybean-Cotton", "Soybean-Corn", "Soybean-Millet", and "Soybean-Sunflower". 

```{r}
# data set of samples
data(samples_mt_6bands)
samples_mt_6bands[1:3,]
```

A `sits tibble` contains data and metadata. The first six columns contain the metadata: spatial and temporal information, label assigned to the sample, and the data cube from where the data has been extracted. The spatial location is given in longitude and latitude coordinates for the "WGS84" ellipsoid. For example, the first sample has been labelled "Cerrado, at location ($-58.5631$, $-13.8844$), and is considered valid for the period (2007-09-14, 2008-08-28). Informing the dates where the label is valid is crucial for correct classification. In this case, the researchers involved in labeling the samples chose to use the agricultural calendar in Brazil, where the spring crop is planted in the months of September and October, and the autumn crop is planted in the months of February and March. For other applications and other countries, the relevant dates will most likely be different from those used in the example. The `time_series` column contains the time series data for each spatiotemporal location. This data is also organized as a tibble, with a column with the dates and the other columns with the values for each spectral band. 

```{r}
# print the first time series records of the first sample
sits_time_series(samples_mt_6bands[1,])[1:3,]
```

The `sits` package provides functions for data manipulation and displaying information for `sits` tibbles. For example, `sits_labels()` shows the labels of the sample set and their frequencies.

```{r}
sits_labels(samples_mt_6bands)
```

In many cases, it is useful to relabel the data set. For example, there may be situations when one wants to use a smaller set of labels, since samples in one label on the original set may not be distinguishable from samples with other labels. We then could use `sits_relabel()`, which requires a conversion list (for details, see `?sits_relabel`).

Given that we have used the tibble data format for the metadata and and the embedded time series, one can use the functions from `dplyr`, `tidyr` and `purrr` packages of the `tidyverse` [@Wickham2017] to process the data. For example, the following code uses `sits_select_bands()` to get a subset of the sample data set with two bands (NDVI and EVI) and then uses the `dplyr::filter()` to select the samples labelled either as "Cerrado" or "Pasture". We can then use the `sits_plot()` to display the time series. Given a small number of samples to display, `sits_plot()` tries to group as many spatial locations together. In the following example, the first 15 samples of  "Cerrado" class refer to the same spatial location in consecutive time periods. For this reason, these samples are plotted together.

```{r cerrado-15, fig.align="center", fig.height=3.1, fig.width=5, fig.cap="Plot of the first 'Cerrado' sample from data set" }
# select NDVI band
samples_ndvi.tb <- sits_select_bands(samples_mt_6bands, ndvi)
# select only samples with Cerrado label
samples_cerrado.tb <-
    dplyr::filter(samples_ndvi.tb, label == "Cerrado")
# plot the first sample
sits_plot(samples_cerrado.tb[1,])
```

For a large number of samples, where the amount of individual plots would be substantial, the default visualization combines all samples together in a single temporal interval (even if they belong to different years). All samples with the same band and label are aligned to a common time interval. This plot is useful to show the spread of values for the time series of each band. The strong red line in the plot shows the median of the values, while the two orange lines are the first and third interquartile ranges. The documentation of `sits_plot()` has more details about the different ways it can display data.

```{r cerrado-all, fig.align="center", fig.height=3.1, fig.width=5, fig.cap="Plot of all Cerrado samples from data set"}
# plot all cerrado samples together
sits_plot(samples_cerrado.tb)
```

# Obtaining time series data

To get a time series in SITS, one has to create a data cube first, as described above. Alternatively, the time series can also be converted from data stored  in the ZOO format [@Zeileis2005]. Users can request one or more time series points from a data cube by using `sits_get_data()`. This function provides a general means of access to image time series. Given data cue, the user provides the latitude and longitude of the desired location, the bands, and the start date and end date of the time series. If the start and end dates are not provided, it retrieves all the available period. The result is a tibble that can be visualized using `sits_plot()`.

```{r, fig.align="center", fig.height=3.1, fig.width=5, fig.cap="NDVI and EVI time series fetched from WTSS service."}
# a point in the transition forest to pasture in Northern MT
# obtain a time series from the WTSS server for this point
series.tb <- sits_get_data(cube      = wtss_cube,
                           longitude = -55.57320, 
                           latitude  = -11.50566,
                           bands     = c("ndvi", "evi"))
sits_plot(series.tb)
```

A useful case is when a set of labelled samples are available to be used as a training data set. In this case, one usually has trusted observations which are labelled and commonly stored in plain text CSV files. Function `sits_get_data()` can get a CSV file path as an argument. The CSV file must provide, for each time series, its latitude and longitude, the start and end dates, and a label associated to a ground sample. 

After importing the samples time series, it is useful to explore the data to see how it is structured and look for its inter-class separability. For example, we can note in the figure above the variability of 400 time series collected from different years and locations. The scattering behavior is intrinsic to remote sensing data. Atmospheric noise, sun angle, interferences on observations or different equipment specifications, as well as the very nature of the climate-land dynamics can be sources of such variability [@Atkinson2012]. One helpful technique to explore such properties is *cluster analysis*. In the following section we present a clustering technique supported by `sits`.

# Filtering techniques 

Satellite image time series generally is contaminated by atmospheric influence, geolocation error, and directional effects [@Lambin2006]. Inter-annual climate variability also changes the phenological cycles of the vegetation, resulting in time series whose periods and intensities do not match on an year to year basis. To make the best use of available satellite data archives, methods for satellite image time series analysis need to deal with  *noisy* and *non-homogeneous* data sets. In this section, we discuss filtering techniques to improve time series data that present missing values or noise.

The literature on satellite image time series have several applications of filtering to correct or smooth vegetation index data. The `sits` have support for Savitzky–Golay (`sits_sgolay()`), Whitaker (`sits_whittaker()`), envelope (`sits_envelope()`) and, the "cloud filter" (`sits_cloud_filter()`) filters. The first two filters are commonly used in the literature, while the remaining two are adapted from other methods and, for our knowledge, its use has not been reported in the literature.

Various somewhat conflicting results have been expressed in relation to the time series filtering techniques for phenology applications. For example, in an investigation of phenological parameter estimation, @Atkinson2012 found that the Whittaker and Fourier transform approaches were preferable to the double logistic and asymmetric Gaussian models. They applied the filters to preprocess MERIS NDVI time series for estimating phenological parameters in India. Comparing the same filters as in the previous work, @Shao2016 found that only Fourier transform and Whittaker techniques improved interclass separability for crop classes and significantly improved overall classification accuracy. The authors used MODIS NDVI time series from the Great Lakes region in North America. @Zhou2016 found that asymmetric Gaussian model outperforms other filters over high latitude boreal biomes, while the Savitzky-Golay model gives the best reconstruction performance in tropical evergreen broadleaf forests. In the remaining biomes, Whittaker gives superior results. The authors compare all previous mentioned filters plus Savitzky-Golay method for noise removal in MODIS NDVI data from sites spread worldwide in different climatological conditions. Many other techniques can be found in applications of satellite image time series such as curve fitting [@Bradley2007], wavelet decomposition [@Sakamoto2005], mean-value iteration, ARMD3-ARMA5, and 4253H [@Hird2009]. Therefore, any comparative analysis of smoothing algorithms depends on the adopted performance measurement.

One of the main uses of time series filtering is to reduce the noise and miss data produced by clouds in tropical areas. The following examples use data produced by the PRODES project [@INPE2017], which detects deforestation in the Brazilian Amazon rain forest through visual interpretation. `sits` provides $617$ samples from a region corresponding to the standard Landsat Path/Row 226/064. This is an area in the East of the Brazilian Pará state. It was chosen because of its huge cloud cover from November to March, which is a significant factor in degrading time series quality. Its NDVI and EVI time series were extracted from a combination of MOD13Q1 and Landsat8 images (to best visualize the effects of each filter, we selected only NDVI time series).

## Savitzky–Golay filter

The Savitzky-Golay filter works by fitting a successive array of $2n+1$ adjacent data points with a $d$-degree polynomial through linear least squares. The central point $i$ of the window array assumes the value of the interpolated polynomial. An equivalent and much faster solution than this convolution procedure is given by the closed expression
$$
{\hat{x}_{i}=\sum _{j=-n}^{n}C_{j}\,x_{i+j}},
$$
where $\hat{x}$ is the the filtered time series, $C_{j}$ are the Savitzky-Golay smoothing coefficients, and $x$ is the original time series.

The coefficients $C_{j}$ depend uniquely on the polynomial degree ($d$) and the length of the window data points (given by parameter $n$). If ${d=0}$, the coefficients are constants ${C_{j}=1/(2n+1)}$ and the Savitzky-Golay filter will be equivalent to moving average filter. When the time series is equally spaced, the coefficients have analytical solution. According to @Madden1978, for ${d\in{}[2,3]}$ each $C_{j}$ smoothing coefficients can be obtained by
$$
C_{j}=\frac{3(3n^2+3n-1-5j^2)}{(2n+3)(2n+1)(2n-1)}.
$$

In general, the Savitzky-Golay filter produces smoother results for a larger value of $n$ and/or a smaller value of $d$ [@Chen2004]. The optimal value for these two parameters can vary from case to case. For example, @Zhou2016 set ${d=2}$ and ${n=10}$. @Hird2009 tests the filter for ${n\in{}[5,6,7]}$ using quadratic polynomial.

`sits` Savitzky-Golay function  The following example shows the effect of Savitsky-Golay filter on the original time series.

```{r, fig.align="center", fig.height=3.1, fig.width=5, fig.cap="Savitzky-Golay filter applied on a one-year NDVI time series."}
# Take NDVI band of the first sample data set
point.tb <- sits_select_bands(prodes_226_064[1,], ndvi)
# apply Savitzky–Golay filter
point_sg.tb <- sits_sgolay(point.tb)
# plot the series
sits_merge(point_sg.tb, point.tb) %>% sits_plot()
```

## Whittaker filter

The Whittaker smoother attempts to fit a curve that represents the raw data, but is penalized if subsequent points vary too much [@Atzberger2011]. The Whittaker filter is a balancing between the residual to the original data and the "smoothness" of the fitted curve. The residual, as measured by the sum of squares of all $n$ time series points deviations, is given by
$$
RSS=\sum_{i}(x_{i} - \hat{x_{i}})^2,
$$
where $x$ and $\hat{x}$ are the original and the filtered time series vectors, respectively. The smoothness is assumed to be the measure of the the sum of the squares of the third order differences of the time series [@Whittaker1922] which is given by
$$
\begin{split}
S\!S\!D = (\hat{x}_4 - 3\hat{x}_3 + 3\hat{x}_2 - \hat{x}_1)^2 + (\hat{x}_5 - 3\hat{x}_4 + 3\hat{x}_3 - \hat{x}_2)^2 \\ + \ldots + (\hat{x}_n - 3\hat{x}_{n-1} + 3\hat{x}_{n-2} - \hat{x}_{n-3})^2.
\end{split}
$$

The filter is obtained by finding a new time series $\hat{x}$ whose points minimize the expression
$$
RSS+\lambda{}S\!S\!D,
$$ 
where $\lambda{}$, a scalar, works as an "smoothing weight" parameter. The minimization can be obtained by differentiating the expression with respect to $\hat{x}$ and equating it to zero. The solution of the resulting linear system of equations gives the filtered time series which, in matrix form, can be expressed as
$$
\hat{x} = ({\rm I} + \lambda {D}^{\intercal} D)^{-1}x,
$$
where ${\rm I}$ is the identity matrix and 
$$
D = \left[\begin{array}{ccccccc}
1 & -3 & 3 & -1 & 0 & 0 &\cdots \\
0 & 1 & -3 & 3 & -1 & 0 &\cdots \\
0 & 0 & 1 & -3 & 3 & -1 & \cdots \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \ddots
\end{array}
\right]
$$
is the third order difference matrix. The Whitakker filter can be a large but sparse optimization problem, as we can note from $D$ matrix. 

Whittaker smoother has been used only recently in satellite image time series investigations. According to @Atzberger2011, the smoother has an advantage over other filtering techniques such as Fourier and wavelets as it does not assume signal periodicity. Moreover, the authors argue that is enables rapid processing of large amounts of data, and handles incomplete time series with missing values. 

In the `sits` package, the Whittaker smoother has two parameters: `lambda` controls the degree of smoothing and `differences` the order of the finite difference penalty. The default values are `lambda = 1` and `differences = 3`. Users should be aware that increasing `lambda` results in much smoother data. When dealing with land use/land cover classes that include both natural vegetation and agriculture, a strong smoothing can reduce the amount of noise in natural vegetation (e.g., forest) time series; however, higher values of `lambda` reduce the information present in agricultural time series, since they reduce the peak values of crop plantations.

The fact that it has only one parameter ($\lambda{}$) facilitates its calibration/comparison process. @Zhou2016 found that $\lambda=15$ gives the best result when compared with $\lambda=2$. Larger values of $\lambda{}$ produces smoother results.

```{r, fig.align="center", fig.height=3.1, fig.width=5, fig.cap="Whittaker smoother filter applied on one-year NDVI time series. The example uses default $\\lambda=1$ parameter."}
# Take EVI band of the first sample data set
point.tb <- sits_select_bands(prodes_226_064[1,], evi)
# apply Whitaker filter
point_whit.tb <- sits_whittaker(point.tb)
# plot the series
sits_merge(point_whit.tb, point.tb) %>% sits_plot()
```

## Envelope filter

This filter can generate a time series corresponding to the superior (inferior) bounding of the input signal. This is accomplished through a convoluting window (odd length) that attributes to the point $i$, in the resulting time series, the maximum (minimum) value of the points in the window. The $i$ point corresponds to the central point of the window. It can be defined as
$$
u_i=\max_{k}{(\{x_{k}:\left|i-k\right|\le{}1\})},
$$
whereas an lower dilation is obtained by
$$
l_i=\min_{k}{(\{x_{k}:\left|i-k\right|\le{}1\})}.
$$
Here, $x$ is the input time series and, $k$ and $i$ are vector indices.

The `sits_envelope()` can combine both maximum and minimum window sequentially. The function can receive a string sequence with `"U"` (for maximization) and `"L"` (for minimization) characters passed to its parameter. A repeated sequence of the same character is equivalent to one operation with a larger window. The sequential operations on the input time series produces the final filtered result that is returned.

The envelope filter can be viewed through mathematical morphology lenses, a very common field in digital image processing [@Haralick1987]. Here the operations of `"U"` and `"L"` corresponds to the *dilation* and *erosion* morphological operators applied to univariate arrays [@Vavra2004]. Furthermore, the compounds operation of *opening* and *closing* can be obtained by `"UL"` and `"LU"`, respectively. This technique has been applied on time series analysis in other fields [@Accardo1997] but, for our knowledge, there is no application in satellite image time series literature.

In the following example we can see an application of `sits_envelope()` function. There, we performs the *opening filtration* and *closing filtration* introduced by @Vavra2004. The correspondent operations sequence are `"ULLULUUL"` and `"LUULULLU"`.

```{r, fig.align="center", fig.height=3.1, fig.width=5, fig.cap="Envelope filter applied on one-year NDVI time series. The examples uses two morfological filters: opening filtration (~.OF) and closing filtration (~.CF)."}
# Take the NDVI band of the first sample data set
point.tb <- sits_select_bands(prodes_226_064[1,], ndvi)
# apply envelope filter (remove downward and upward noises)
point_env1.tb <- 
    sits_envelope(point.tb,
                  "ULLULUUL",
                  bands_suffix = "OF")
point_env2.tb <- 
    sits_envelope(point.tb,
                  "LUULULLU",
                  bands_suffix = "CF")
# plot the series
sits_merge(point_env1.tb, point_env2.tb) %>%
    sits_merge(point.tb) %>%
    sits_plot()
```

## ARIMA filter for cloud removal

The cloud filter makes use of the well known autoregressive integrated moving average (ARIMA) model. The algorithm looks to the first order difference time series for points where the value is above a certain threshold. This procedure selects only those points with large variations in the original time series, probably associated with noise. Finally, these points are replaced by the ARIMA correspondent values.

The parameters of the ARIMA model can be set by the user. Please see `arima` for the detailed description of parameters $p$, $d$, and $q$.

```{r, fig.align="center", fig.height=3.1, fig.width=5}
# Take the NDVI band of the first sample data set
point.tb <- sits_select_bands(prodes_226_064[1,], ndvi)
# apply ARIMA filter
point_cf.tb <- sits_ndvi_arima_filter(point.tb, apply_whit = FALSE)
# plot the series
sits_merge(point_cf.tb, point.tb) %>% sits_plot()
```

#Pre-processing for sample quality control

One of the key challenges when using samples to train machine learning classification models is assessing their quality. Noisy and imperfect training samples can have a negative effect on classification performance [@Frenay2013].  Therefore, it is useful to apply pre-processing methods to improve the quality of the samples and to remove those that might have been wrongly labeled or that have low discriminatory power. Representative samples lead to good classification maps. `sits` provides support for two clustering methods to test sample quality: (a) Agglomerative Hierarchical Clustering (AHC); (b)  Self-organizing Maps (SOM).

## Hierachical clustering

Cluster analysis has been used for many purposes in satellite image time series literature ranging from unsupervised classification [@Petitjean2011], and pattern detection [@Romani2011]. Here, we are interested in the second use of clustering, using it as a way to improve training data to feed machine learning classification models. In this regard, cluster analysis can assist the identification of structural *time series patterns* and anomalous samples [@Romani2011], [@Chandola2009]. 

Agglomerative hierarchical clustering (AHC) is a family of methods that groups elements using a distance function to associate a real value to a pair of elements. From this distance measure, we can compute the dissimilarity between any two elements from a data set. Depending on the distance functions and linkage criteria, the algorithm decides which two clusters are merged at each iteration. AHC approach is suitable for the purposes of samples data exploration due to its visualization power and ease of use [@Keogh2003]. Moreover, AHC does not require a predefined number of clusters as an initial parameter. This is an important feature in satellite image time series clustering since defining the number of clusters present in a set of multi-attribute time series is not straightforward [@Aghabozorgi2015].

The main result of AHC method is a *dendrogram*. It is the ultrametric relation formed by the successive merges in the hierarchical process that can be represented by a tree. Dendrograms are quite useful to decide the number of clusters to partition the data. It shows the height where each merging happens, which corresponds to the minimum distance between two clusters defined by a *linkage criterion*. The most common linkage criteria are: *single-linkage*, *complete-linkage*, *average-linkage*, and *Ward-linkage*. Complete-linkage prioritizes the within-cluster dissimilarities, producing clusters with shorter distance samples. Complete-linkage clustering can be sensitive to outliers, which can increase the resulting intracluster data variance. As an alternative, Ward proposes a criteria to minimize the data variance by means of either *sum-of-squares* or *sum-of-squares-error* [@Ward1963]. Ward's intuition is that clusters of multivariate observations, such as time series, should be approximately elliptical in shape [@Hennig2015]. In `sits`, a dendrogram can be generated by `sits_dendrogram()`. The following codes illustrate how to create, visualize, and cut a dendrogram (for details, see `?sits_dendrogram()`).


After creating a dendrogram, an important question emerges: *where to cut the dendrogram?* The answer depends on what are the purposes of the cluster analysis [@Hennig2015]. If one is interested in an unsupervised classification, it is common to use *internal validity indices*, such as silhouettes [@Rousseeuw1987], to help determine the best number of clusters. However, if one is interested in understanding the structure of a labeled data set, or in the identifying sample anomalies, as we are here, one can use *external validity indices* to assist the semi-supervised procedure in order to achieve the optimal correspondence between clusters and classes partitions. In this regard, we need to balance two objectives: get clusters as large as possible, and get clusters as homogeneous as possible with respect to their known classes. To help this process, `sits` provides `sits_dendro_bestcut()` function that computes an external validity index *Adjusted Rand Index* (ARI) for a series of different number of generated clusters. This function returns the height where the cut of the dendrogram maximizes the index.

In this example, the height optimizes the ARI and generates $6$ clusters. The ARI considers any pair of distinct samples and computes the following counts:
(a) the number of distinct pairs whose samples have the same label and are in the same cluster;
(b) the number of distinct pairs whose samples have the same label and are in different clusters;
(c) the number of distinct pairs whose samples have different labels and are in the same cluster; and
(d) the number of distinct pairs whose samples have the different labels and are in different clusters.
Here, $a$ and $d$ consist in all agreements, and $b$ and $c$ all disagreements. The ARI is obtained by:

$$
ARI=\frac{a+d-E}{a+d+b+c-E},
$$
where $E$ is the expected agreement, a random chance correction calculated by 
$$
E=(a+b)(b+c)+(c+d)(b+d).
$$ 

Unlike other validity indexes such as Jaccard (${J=a/(a+b+c)}$), Fowlkes-Mallows (${FM=a/(a^2+a(b+c)+bc)^{1/2}}$), and Rand (the same as ARI without the $E$ adjustment) indices, ARI is more appropriate either when the number of clusters is outweighed by the number of labels (and *vice versa*) or when the amount of samples in labels and clusters is imbalanced [@Hubert1985], which is usually the case.

```{r dendrogram, cache=TRUE, fig.align="center", fig.height=4.1, fig.width=5}
# take a set of patterns for 2 classes
# create a dendrogram object, plot, and get the optimal cluster based on ARI index
clusters.tb <- sits_cluster(cerrado_2classes, bands = c("ndvi", "evi"))

# show clusters samples frequency
sits_cluster_frequency(clusters.tb)
```

Note in this example that almost all clusters has a predominance of either "Cerrado" or "Pasture" classes with the exception of cluster $3$. The contingency table plotted by `sits_cluster_frequency()` shows how the samples are distributed across the clusters and helps to identify two kinds of confusions. The first is relative to those small amount of samples in clusters dominated by another class (*e.g.* clusters $1$, $2$, $4$, $5$, and $6$), while the second is relative to those samples in non-dominated clusters (*e.g.* cluster $3$). These confusions can be an indication of samples with poor quality, an inadequacy of selected parameters for cluster analysis, or even a natural confusion due to the inherent variability of the land classes.

The result of the `sits_cluster` operation is a `sits_tibble` with one additional column, called "cluster". Thus, it is possible to remove clusters with mixed classes using standard `R` such as those in the `dplyr` package. In the example above, removing cluster $3$ can be done using the `dplyr::filter` function.

```{r}
# remove cluster 3 from the samples
clusters_new.tb <- dplyr::filter(clusters.tb, cluster != 3)

# show new clusters samples frequency
sits_cluster_frequency(clusters_new.tb)
```

The resulting clusters still contained mixed labels, possibly resulting from outliers. In this case, users may want to remove the outliers and leave only the most frequent class. To do this, one can use `sits_cluster_clean()`, which removes all minority samples, as shown below..

```{r}
# clear clusters, leaving only the majority class in each cluster
cleaned.tb <- sits_cluster_clean(clusters.tb)
# show clusters samples frequency
sits_cluster_frequency(cleaned.tb)
```

## Self-organizing maps

Self-organizing maps is a dimensionality reduction technique [@Kohonen2001]. High-dimensional data is mapped into two dimensions, keeping the topological relations between data patterns. This allows users to visualize and assess the structure of the input dataset. For quality control of the training data, `sits` uses a SOM clustering method, whose input is a dataset of time series samples. Package kohonen [@Wehrens2018] is used as basis to run SOM.

The output layer comprises a 2D grid of neurons, each associated to a weight vector of the same dimension as the input space. The 2D grid of neurons is initialized randomly. Then, for each time series sample, the algorithm finds the 2D grid neuron which has the smallest distance, based on its weight vector. After the match, the neuron's weight vector and those of its neighbors are then updated. After all training samples are associated with neurons, each neuron is labeled using a majority vote, taking the most frequent class from the samples associated with it. In this way, SOM splits the output space into regions. It is expected that good quality samples of each class should be close together in the resulting map. This allows using SOM to detect and remove outliers. 

To increase the reliability of quality control procedures, the SOM clustering can be  executed several times. Using this iterative procedure, we computed the probability that a sample belongs to each of the resulting clusters. From these probabilities, we analyzed the separability of samples with similar phenological patterns and decided which samples were to be discarded.

For each neuron $j$, SOM uses a vector of weights, $w_j=[w_{j1},\ldots,w_{jn} ]$,that has the same dimension of the input vector of time series samples $x(t)=[x(t)_1,\ldots,x(t)_n ]$, where $n$ is the time series dimension. In the begining, the neurons are randomly initialized.
At each training step, a time series sample $x(t)$ is presented to the network in order to find the neuron whose weight vector has the smaller distance to the sample. Distance metrics $D_{j}$ compute the distance between input sample $x(t)$ and the weight vector of a neuron. This is computed for each neuron $j$ in the output layer. The most commonly used metric is Euclidean distance, shown below.

$$
D_{j}=\sum_{i=1}^{n}{\sqrt{(x(t)_i{-}w_{ji})^{2}}}.
$$

The neuron that contains the shortest distance is declared the winner neuron, defined here as ${b}$, also know as Best Matching Unit (BMU):


$$
 d_{b}= min \left\{D_1,\ldots, D_J \right\}.
$$

Once the BMU is selected for a given sample, its neighborhood must be updated. The weight of each neighbour is adjusted according to  the similarity with input vector.
The equation for updating the weight vector is given by:

$$
 w_{j}(t{+}1)=  w_{j}(t){+}\alpha(t)* h_{b,j}(t) [x(t)_i{-}w_{j}(t)],
$$
where $\alpha(t)$ is the learning rate, which must be set as $0<\alpha(t)<1$, and $h_{b,j}(t)$ is a neighbourhood function.

After producing the SOM map, it is necessary to separate the samples in groups in order to label the neurons. In this step, each neuron is labeled using majority vote from  the labels of its samples. As a result of this, a sample belonging to a neuron with a different label will have its own label updated to the label of the neuron. When the neuron has no samples, it will receive the label "Noclass". Clusters are then composed by all neurons that belong to the same label.

```{r}
data("samples_mt_9classes", package = "sits")

```
```{r}
data.tb <- samples_mt_9classes
```


```{r}
# clustering time series using SOM
som_cluster.tb <-
    sits::sits_cluster_som(
        data.tb,
        grid_xdim = 25,
        grid_ydim = 25,
        rlen = 100,
        distance = "euclidean",
        mode = "pbatch",
        iterations = 2

    )
# return a tibble with all samples and which cluster it belongs

```

The function `sits_plot_som()` plots the kohonen map. The neurons are labelled using the majority voting.
```{r}
sits_plot_som(som_cluster.tb)
```

The process of clustering  with SOM can be performed several times to evaluate the reliability of each sample. This can be executed by `sits_cluster_som()`. This procedure returns a list. This list returns a tibble with statistics of each sample and neuron and its neighbours in each iteration and a summarized data.frame that contains the probabilities of each sample belonging to a cluster.


For each sample, a tibble contains a sample identifier, its original label, the label of the neuron, the id of the neuron, and the probability of the sample belonging to the neuron's label and the number of the iteration that was obtained the information about the cluster. The field probability was computed  through bayesian filter using the probability of the samples belongs to a cluster and its neighbourhood on the kononen map.
```{r}
statistics_about_som_cluster <- som_cluster.tb$statistics_samples
detail_samples.tb <- statistics_about_som_cluster$samples
detail_samples.tb
```

For each neuron, a tibble contains a neuron identifier,  the label of the neuron, the frequency and labels of each neuron neighbor, the number of neigbours that a neuron contain and the number of the iteration that was obtained each information.

```{r}
detail_neurons.tb <- statistics_about_som_cluster$neuron
detail_neurons.tb
```

Another element returned by `sits_cluster_som` about the statistics is a contingency table, that contains the probabilities of each sample belonging to a cluster. It is a tibble where each row contains a sample identifier, its original label, the label of the neuron, and the probability of the sample belonging to the neuron's label, according to the frequency in the different repetitions of SOM. Since we have the probability calculated through the Bayesian filter for each sample in each iteration, the median is calculated to obtain the total probability of sample $s_k$ belonging to a cluster. From these results, the user can make decisions to improve classification, such as manually removing samples of dataset. 
```{r}
summarized_sample_cluster.tb <- statistics_about_som_cluster$cluster_sample_probability
summarized_sample_cluster.tb
```

From the statistics obtained from function `sits_cluster_som`, a new dataset can be generated removing the samples with bad quality.
The function `sits_clean_samples_som()` remove the bad samples based on the probability a samples belongs to a cluster. 

```{r}
new_samples.tb <- sits_clean_samples_som(som_cluster.tb)
new_samples.tb
```

To verify the quality of the clusters generated by SOM, a confusion matrix, the overall accuracy and the statistics about each class is evaluated.  
```{r}
cluster_overall <- sits_evaluate_cluster(som_cluster.tb)
cluster_overall$confusion_matrix
```
Besides that, a table showing how a table showing how pure the cluster is computed.
```{r}
cluster_overall$mixture_cluster
```

Finally, a graphic showing the mixture within each cluster is presented.
```{r}
sits_plot_cluster_info(cluster_overall, "Confusion by cluster")
```
A class can be represent by several patterns. The function `sits_evaluate_som_subgroups()` takes a cluster and estimates how many subgroups it contains, using the hierachical clustering a cluster is partioned in several subgorups. A tibble of samples with new column showing the subgroups is returned.

```{r}
subgroups <- sits_evaluate_som_subgroups(som_cluster.tb)
subgroups$samples_subgroups.tb

```
The function `sits_plot_subgroups()`is used to visualize the signature of each subgroup.

```{r}
sits_plot_subgroups(subgroups, cluster = "Soy_Corn", band = "ndvi")
```



# Machine learning classification

There has been much recent interest in using classifiers such as support vector machines \citep{Mountrakis2011} and random forests \citep{Belgiu2016} for remote sensing images. Most often, researchers use a \emph{space-first, time-later} approach, in which the dimension of the decision space is limited to the number of spectral bands or their transformations. Sometimes, the decision space is extended with temporal attributes.  To do this, researchers filter the raw data to get smoother time series \citep{Brown2013, Kastens2017}. Then, using software such as TIMESAT \citep{Jonsson2004}, they derive a small set of phenological parameters from vegetation indexes, like the beginning, peak, and length of the growing season \citep{Estel2015, Pelletier2016}. 

`sits` has support for a variety of machine learning techniques: linear discriminant analysis, quadratic discriminant analysis, multinomial logistic regression, random forests, support vector machines, and deep learning. In a recent review of machine learning methods to classify remote sensing data [@Maxwell2018], the authors note that many factors influence the performance of these classifiers, including the size and quality of the training dataset, the dimension of the feature space, and the choice of the parameters. We support both \emph{space-first, time-later} and \emph{time-first, space-later} approaches. Therefore, the `sits` package provides functionality to explore the full depth of satellite image time series data. 

When used in \emph{time-first, space-later} approache, `sits` treats time series as a feature vector. To be consistent, the procedure aligns all time series from different years by its time proximity considering an given cropping schedule. Once aligned, the feature vector is formed by all pixel "bands". The idea is to have as many temporal attributes as possible, increasing the dimension of the classification space. In this scenario, statistical learning models are the natural candidates to deal with high-dimensional data: learning to distinguish all land cover and land use classes from trusted samples exemplars (the training data) to infer classes of a larger data set. 

## Support Vector Machine

Given a multidimensional dataset, the Support Vector Machine (SVM) method finds an optimal separation hyperplane that minimizes misclassifications [@Cortes1995]. Hyperplanes are linear ${(p-1)}$-dimensional boundaries and define linear partitions in the feature space. The solution for the hyperplane coefficients depends only on those samples that violates the maximum margin criteria, the so-called *support vectors*. All other points far away from the hyperplane does not exert any influence on the hyperplane coefficients which let SVM less sensitive to outliers.

For data that is not linearly separable, SVM includes kernel functions that map the original feature space into a higher dimensional space,  providing nonlinear boundaries to the original feature space. In this manner, the new classification model, despite having a linear boundary on the enlarged feature space, generally translates its hyperplane to a nonlinear boundaries in the original attribute space. The use of kernels are an efficient computational strategy to produce nonlinear boundaries in the input attribute space an hence can improve training-class separation. SVM is one of the most widely used algorithms in machine learning applications and has been widely applied to classify remote sensing data [@Mountrakis2011].

In `sits`, SVM is the default machine learning model. As a wrapper of `e1071` R package that uses the `LIBSVM` implementation [@Chang2011], `sits` adopts the *one-against-one* method for multiclass classification. For a $q$ class problem, this method creates ${q(q-1)/2}$ SVM binary models, one for each class pair combination and tests any unknown input vectors throughout all those models. The overall result is computed by a voting scheme. The following example illustrate how to classify an individual sample (in this case a series of 16 one-year NDVI time series from the same location). We used the NDVI time series from Mato Grosso Brazilian state as a training data set.

```{r, fig.align="center", fig.height=3.4, fig.width=5.5, fig.cap="SVM classification of a $16$ years time series. The location (latitude, longitude) shown at the top of the graph is in geographic coordinate system (WGS84 {\\it datum})."}
# Retrieve the set of samples (provided by EMBRAPA) from the 
# Mato Grosso region for train the SVM model
data(samples_mt_ndvi)
# train a machine learning model using SVM
svm_model <- sits_train(samples_mt_ndvi,
                        sits_svm(kernel = "radial",
                                 cost = 10))
# get a point to be classified
data(point_ndvi)
# Classify using SVM model
class.tb <- sits_classify(point_ndvi, svm_model)
sits_plot(class.tb)
```

## Random forest

Random forest uses the idea *decision tree* as its base model. It combines many decision trees via *bootstrap* procedure and *stochastic feature selection*, developing a population of somewhat uncorrelated base models. The final classification model is obtained by a majority voting schema. This procedure decreases the classification variance, improving prediction of individual decision trees. 

Random forest training process is essentially nondeterministic. It starts by growing trees through repeatedly random sampling-with-replacement the observations set. At each growing tree, the random forest considers only a fraction of the original attributes to decide where to split a node, according to a *purity criterion*. This decreases the correlation among trees and improves the prediction performance. The most used impurity criterion are *Gini*, *cross-entropy*, and *misclassification error*. The splitting process continues until the tree reaches some given minimum nodes size or a minimum impurity index value.

The random forest classification performance can vary according to the tunning of the model parameters. The main random forest parameters are the number of attributes sampled as candidates at each split, the number of decision trees to grow, the minimum node size, and the sample fraction to be drawn at each bootstrap iteration. 

<!--
% Lower values of $m$ tends to grow taller and uncorrelated decision trees, but decreases the model classification accuracy. The parameter $b$, the number of trees composing the random forest ensemble, is connected with the variance reduction of the model. High values of $b$ can reduce the variance of the model up to a certain level (dependent on the observations data set) but reduces the model performance. Lower values of $N_{\!m\!i\!n}$ parameter can introduce model overfitting but may increase the classification accuracy. Finally, lower values of $\lambda$ tends to increase the model variance but can underrepresent the sample universe reducing its accuracy.
-->

```{r, fig.align="center", fig.height=3.4, fig.width=5.5, fig.cap="Random forest classification of a $16$ years time series. The location (latitude, longitude) shown at the top of the graph are in geographic coordinate system  (WGS84 {\\it datum})."}
# Retrieve the set of samples (provided by EMBRAPA) from the 
# Mato Grosso region for train the Random Forest model.
data(samples_mt_ndvi)
# train a machine learning model using random forest
rfor_model <- sits_train(samples_mt_ndvi, sits_rfor(num.trees = 1000))
# get a point to be classified
data(point_ndvi)
# Classify using Random Forest model
class.tb <- sits_classify(point_ndvi, rfor_model)
sits_plot(class.tb)
```

## Deep learning methods 

In `sits`, the interface to deep learning models is done using the `keras` package [@Chollet2018]. This package implements different deep learning techniques. Currently, `sits` provides access to deep feedforward networks. Also called feedforward neural networks, or multilayer perceptrons (MLPs), these are the quintessential deep learning models. The goal of a feedforward network is to approximate some function $f$. For example, for a classifier $y =f(x)$ maps an input $x$ to a category $y$. A feedforward network defines a mapping $y = f(x;\theta)$ and learns the value of the parameters $\theta$ that result in the best function approximation. These models are called feedforward because information flows through the function being evaluated from $x$, through the intermediate computations used to define $f$, and finally to the output $y$. There are no feedback connections in which outputs of the model are fed back into itself [@Goodfellow2016].

Specifying a MLP requires some work on customization, which requires some amount of trial-and-error by the user, since there is no proven model for classification of satellite image time series. The most important decision is the number of layers in the model. Initial tests indicate that 3 to 5 layers are enough to produce good results. The choice of the number of layers depends on the inherent separability of the data set to be classified. For data sets where the classes have different signatures, a shallow model (with 3 layers) may provide appropriate responses. More complex situations require models of deeper hierarchy. The user should be aware that some models with many hidden layers may take a long time to train and may not be able to converge. The suggestion is to start with 3 layers and test different options of number of neurons per layer, before increasing the number of layers.

Three other important parameters for an MLP are: (a) the activation function; (b) the optimization method; (c) the dropout rate. The activation function  the activation function of a node defines the output of that node given an input or set of inputs. Following standard practices [@Goodfellow2016], we recommend the use of the "relu" and "elu" functions. The optimization method is a crucial choice, and the most common choices are gradient descent algorithm. These methods aim to maximize an objective function by updating the parameters in the opposite direction of the gradient of the objective function [@Ruder2016]. Based on experience with image time series, we recommend that users start by using the default method provided by `sits`, which is the `optimizer_adam` method. Please refer to the `keras` package documentation for more information.

The dropout rates have a huge impact on the performance of deep learning classifiers. Dropout is a technique for randomly  dropping  units  from  the  neural network during training [@Srivastava2014]. By randomly discarding some neurons, dropout reduces overfitting. It is a counter-intuitive idea that works well. Since the purpose of a cascade of neural nets is to improve learning as more data is acquired, discarding some of these neurons may seem a waste of resources. In fact, as experience has shown [@Goodfellow2016], this procedures prevents an early convergence of the optimization to a local minimum. Thus, in practice, dropout rates between 50% and 20% are recommended for each layer. 

In the following example, we classify the same data set using a simple example of the `deep learning` method, for fast processing: (a) Two layers with 512 neurons each; (b) Using the 'elu' activation function and 'optimizer_adam'; (c) dropout rates of 40% and 30% for the layers. 

```{r, fig.align="center", fig.height=3.4, fig.width=5.5, fig.cap="Deep learning classification of a $16$ year time series. The location (latitude, longitude) shown at the top of the graph are in geographic coordinate system  (WGS84 {\\it datum})."}
# Retrieve the set of samples (provided by EMBRAPA) from the 
# Mato Grosso region for train the  model.
data(samples_mt_ndvi)
# train a machine learning model using deep learning
train_dl <- sits_deeplearning(
    units            = c(512, 512),
    activation       = "elu",
    dropout_rates    = c(0.40, 0.30),
    optimizer        = keras::optimizer_adam(lr = 0.001),
    epochs           = 50,
    batch_size       = 128,
    validation_split = 0.2)

dl_model <- sits_train(samples_mt_ndvi, train_dl)
# get a point to be classified
data(point_ndvi)
class.tb <- sits_classify(point_ndvi, dl_model)
sits_plot(class.tb)
```

# Validation techniques

Validation is a process undertaken on models to estimate some error associated with them, and hence has been used widely in different scientific disciplines. Here, we are interested in estimating the prediction error associated to some model. For this purpose, we concentrate on the *cross-validation* approach, probably the most used validation technique [@Hastie2009].

To be sure, cross-validation estimates the expected prediction error. It uses part of the available samples to fit the classification model, and a different part to test it. The so-called *k-fold* validation, we split the data into $k$ partitions with approximately the same size and proceed by fitting the model and testing it $k$ times. At each step, we take one distinct partition for test and the remaining ${k-1}$ for training the model, and calculate its prediction error for classifying the test partition. A simple average gives us an estimation of the expected prediction error. 

A natural question that arises is: *how good is this estimation?* According to @Hastie2009, there is a bias-variance trade-off in choice of $k$. If $k$ is set to the number of samples, we obtain the so-called *leave-one-out* validation, the estimator gives a low bias for the true expected error, but produces a high variance expectation. This can be computational expensive as it requires the same number of fitting process as the number of samples. On the other hand, if we choose ${k=2}$, we get a high biased expected prediction error estimation that overestimates the true prediction error, but has a low variance. The recommended choices of $k$ are $5$ or $10$ [@Hastie2009], which somewhat overestimates the true prediction error.

`sits_kfold_validate()` gives support the k-fold validation in `sits`. The following code gives an example on how to proceed a k-fold cross-validation in the package. It perform a five-fold validation using SVM classification model as a default classifier. We can see in the output text the corresponding confusion matrix and the accuracy statistics (overall and by class).

```{r}
# read a set of samples
data(cerrado_2classes)

# perform a five fold validation with the 
# SVM machine learning method using default parameters
prediction.mx <- 
    sits_kfold_validate(cerrado_2classes, 
                        folds = 5,
                        ml_method = sits_svm())
# prints the output confusion matrix and statistics 
sits_conf_matrix(prediction.mx)
```

# Cube classification

The continuous observation of the Earth surface provided by orbital sensors is unprecedented in history. Just for the sake of illustration, a unique tile from MOD13Q1 product, a square of $4800$ pixels provided every 16 days since February 2000 takes around $18$GB of uncompressed data to store only one band or vegetation index. This data deluge puts the field into a big data era and imposes challenges to design and build technologies that allow the Earth observation community to analyse those data sets [@Gilberto2017]. 

To classify a data cube, use the function `sits_classify()` as described below. This function works both with cubes built from raster bricks and those built with services such as "EOCUBES". The classification algorithms allows users to choose how many process will run the task in parallel, and also the size of each data chunk to be consumed at each iteration. This strategy enables `sits` to work on average desktop computers without depleting all computational resources. The code bellow illustrates how to classify a small raster brick image that accompany the package.

## Steps for cube classification 

Once a data cube which has associated files is defined, the steps for classification are:

1. Select a set of training samples.
2. Train a machine learning model
3. Classify the data cubes using the model, producing a data cube with class probabilities.
4. Label the cube with probabilities, including data smoothing if desired.

## Adjustments for improved performance 

To reduce processing time, it is necessary to adjust `sits_classify()` according to the capabilities of the server. The package tries to keep memory use to a minimum, performing garbage collection to free memory as often as possible. Nevertheless, there is an inevitable trade-off between computing time, memory use, and I/O operations. The best trade-off has to be determined by the user, considering issues such disk read speed, number of cores in the server, and CPU performance.

The first parameter is `memsize`. It controls the size of the main memory (in GBytes) to be used for classification. The user must specify how much free memory will be available. The second factor controlling performance of raster classification is `multicores`. Once a block of data is read from disk into main memory, it is split into different cores, as specified by the user. In general, the more cores are assigned to classification, the faster the result will be. However, there are overheads in switching time, especially when the server has other processes running.

Based on current experience, the classification of a MODIS tile (4800 x 4800) with four bands and 400 time instances, covering 15 years of data, using SVM with a training data set of about 10,000 samples, takes about 24 hours using 20 cores and a memory size of 60 GB, in a server with 2.4GHz Xeon CPU and 96 GB of memory to produce the yearly classification maps.

```{r, fig.align="center", fig.height=3.4, fig.width=4.1, fig.cap="Image (${11\\times14}$ pixels) classified using SVM. The image coordinates ({\\it meters}) shown at vertical and horizontal axis are in MODIS sinusoidal projection."}
# Retrieve the set of samples for the Mato Grosso region 
data(samples_mt_ndvi)

# build a machine learning model for this area
svm_model <- sits_train(samples_mt_ndvi, sits_svm())

# read a raster file and put it into a vector
file <- system.file("extdata/raster/mod13q1/sinop-crop-ndvi.tif", 
                    package = "sits")

# define the timeline
data("timeline_modis_392")

# create a raster metadata file based on the 
# information about the files
raster_cube <-  sits_cube(name = "Sinop-crop", timeline = timeline_modis_392,
              bands    = "ndvi", files    = file)

# classify the raster file, generating a probability file
probs_cube <- sits_classify(raster_cube, ml_model = svm_model, memsize = 2, multicores = 1)

# label the probability file (by default selecting the class with higher probability)
label_cube <- sits_label_classification(probs_cube)

# plot the first raster object with a selected color pallete
# make a title, define the colors and the labels)
sits_plot_raster(label_cube, time = 1, title = "SINOP-MT - 2000/2001")
```

The classified files can also be visualized with applications such as QGIS. Note that we create two coverage `tibbles` with metadata information, one for the input data and other for the output.  To create the input data, we need a `timeline` that matches the input images of the raster brick. Once created, the coverage can be used either to retrieve time series data from the raster bricks using `sits_get_data()` or to do the raster classification by calling the function `sits_classify_raster()`. The machine learning model and the training data to be used are passed by the arguments `sits_svm()` (default) and `samples_mt_ndvi` parameters. The classification result is stored as a set of files beginning with `file` prefix.

```{r, include=FALSE}
# remove all files
file.remove(unlist(probs_cube$files))
```
# Smoothing of raster data after classification

One of the well-established methods in remote sensing image analysis is to combine pixel-based classification methods with a spatial post-processing method to remove outliers and misclassified pixels. Methods proposed in the literature include modal filters [@Ghimire2010] and probabilistic relaxation [@Gong1989]. Our method used Bayesian smoothing to reclassify the pixels because machine learning methods assign class probabilities to each pixel. 

Most applications of machine learning methods for image classification use only the categorical result of the classifier which is the most probable class. The proposed method uses all class probabilities to compute our confidence in the result. For each pixel $i$, the odds of the classification for class $k$ is $o_{i,k} = p_{i,k} / (1-p_{i,k})$ where $p_{i,k}$ is the probability of class $k$. We have more confidence in pixels with higher odds since their class assignment is stronger. However, there are situations, such as a border or mixed pixels, where the odds of different classes are similar in magnitude. We took this as a case of low confidence in the classification result. To assess these situations, we followed well-established Bayesian smoothing methods [@Cressie1995]: borrow strength from the neighbors and thus reduced the variance of the estimated class for each pixel.

More formally, consider the transformation $l_{i,k} = \log [o_{i,k}]$ which measures the logit (log of the odds) associated to classifying the pixel $i$ as being of class $k$. The support of $l_{i,k}$ is $\mathbb{R}$. Let $V_{i}$ be a spatial neighbourhood for pixel $i$. We considered a Bayesian normal distribution update for the value $l_{i,k}$ based on the neighbourhood, assuming independence between the classes. In this way, the update be performed for each class $k$ at a time. Consider that $l_{i,k}$ follows a normal distribution, $N(\mu_{i,k}, \sigma^2_{k})$, with parameters $\mu_{i,k}$ and $\sigma^2_{k}$. 
In a Bayesian context, the likelihood function is given by
$$
l_{i,k} | \mu_{i,k} \sim N(\mu_{i,k}, \sigma^2_{k})
$$
We assumed that all pixels in the neighbourhood follow the same normal prior for the parameter $\mu_{i,k}$ with parameters $m_{i,k}$ and $s^2_{i,k}$, so that
$$
\mu_{i,k} \sim N(m_{i,k}, s^2_{i,k}).
$$
The Bayesian smoothing procedure then comprises conditionally estimating the value of $\mu_{i,k}$, which is the updated value of the logit of class probability for class $k$ of pixel $i$. We used $\sigma^2_{k}$ as a hyper-parameter to control the level of smoothness. Based on Bayesian statistics, the value of conditional mean for a normal distribution is given by:
$$
{E}[\mu_{i,k} | l_{i,k}] =
\frac{m_{i,t} \times \sigma^2_{k} + 
l_{i,k} \times s^2_{i,k}}{ \sigma^2_{k} +s^2_{i,k}}
$$
which can also be expressed as
$$
{E}[\mu_{i,k} | l_{i,k}] =
\Biggl [ \frac{s^2_{i,k}}{\sigma^2_{k} +s^2_{i,k}} \Biggr ] \times
l_{i,k} +
\Biggl [ \frac{\sigma^2_{k}}{\sigma^2_{k} +s^2_{i,k}} \Biggr ] \times m_{i,k}
$$
In the above equation, the parameter $m_{i,k}$ is the local mean of the probability distribution of values for class $k$ and $s^2_{i,k}$ is the local variance. We estimated local means and variances by considering the neighboring pixels in space. Let $\#(V_{i})$ be the number of elements in $V_{i}$. We then have for the local mean:

$$
m_{i,k} = \frac{\displaystyle\sum_{j \in V_{i}} l_{j,k}}{\#(V_{i})},
$$
and for the local standard deviation:
$$
s^2_{i,k} = \frac{\displaystyle\sum_{j \in V_{i}} [l_{j,k} - m_{i,k}]^2}{\#(V_{i})-1}.
$$

The updated value for the class probability of the pixel is a weighted average between the original logit value $l_{i,k}$ and the mean of the class logits $m_{i,k}$ for the neighboring pixels. When the local class variance of the neighbors $s^2_{i,k}$ is high, our confidence on the influence of the neighbors is low, and the smoothing algorithm gives more weight to the original pixel value $l_{i,k}$. When the local class variance $s^2_{i,k}$ decreases relative to the smoothness factor $\sigma^2_k$, then our confidence on the influence of the neighborhood increases. The smoothing procedure will be most relevant in situations where the original classification odds ratio is low, showing a low level of separability between classes. In these cases, the updated values of the classes will be influenced by the local class variances. 

The parameter $\sigma^2_k$ sets the level of smoothness. If $\sigma^2_k$ is zero, the smoothed value ${E}[\mu_{i,,k} | l_{i,k}]$ is equal to the pixel value $l_{i,k}$. Higher values of $\sigma^2_k$ will cause the assignment of the local mean to the pixel updated probability. In practice, $\sigma^2_k$ is a user-controlled parameter that will be set by users based on their knowledge of the region to be classified. In our case, after some classification tests, we decided to set the parameters $V$ as the Moore neighborhood where each pixel is connected to all those pixels with Chebyshev distance of $1$, and $\sigma^2_k=10$ for all $k$. This level of smoothness showed the best performance in the technical validation.


# Final remarks

Current approaches to image time series analysis still use limited number of attributes. A common approach is deriving a small set of phenological parameters from vegetation indices, like beginning, peak, and length of growing season [@Brown2013], [@Kastens2017], [@Estel2015], [@Pelletier2016]. These phenological parameters are then fed in specialized classifiers such as TIMESAT [@Jonsson2004]. These approaches do not use the power of advanced statistical learning techniques to work on high-dimensional spaces with big training data sets [@James2013].

Package `sits` can use the full depth of satellite image time series to create larger dimensional spaces. We tested different methods of extracting attributes from time series data, including those reported by @Pelletier2016 and @Kastens2017. Our conclusion is that part of the information in raw time series is lost after filtering. Thus, the method we developed uses all the data available in the time series samples. The idea is to have as many temporal attributes as possible, increasing the dimension of the classification space. Our experiments found out that modern statistical models such as support vector machines, and random forests perform better in high-dimensional spaces than in lower dimensional ones. 

# Acknowledgements

The authors would like to thank all the researchers that provided data samples used in the examples: Alexandre Coutinho, Julio Esquerdo and Joao Antunes (Brazilian Agricultural Research Agency, Brazil) who provided ground samples for "soybean-fallow", "fallow-cotton", "soybean-cotton", "soybean-corn", "soybean-millet", "soybean-sunflower", and "pasture" classes; Rodrigo Bergotti (National Institute for Space Research, Brazil) who provided samples for "cerrado" and "forest" classes; and Damien Arvor (Rennes University, France) who provided ground samples for "soybean-fallow" class. 

This work was partially funded by the São Paulo Research Foundation (FAPESP) through eScience Program grant 2014/08398-6. We thank the Coordination for the Improvement of Higher Education Personnel (CAPES) and National Council for Scientific and Technological Development (CNPq) grants 312151/2014-4 (GC) and 140684/2016-6 (RS). We thank Ricardo Cartaxo, Lúbia Vinhas, and Karine Ferreira who provided insight and expertise to support this paper.

This work has also been supported  by the International Climate Initiative of the Germany Federal Ministry for the Environment, Nature Conservation, Building and Nuclear Safety under Grant Agreement 17-III-084-Global-A-RESTORE+ (``RESTORE+: Addressing Landscape Restoration on Degraded Land in Indonesia and Brazil''). 

<!--
# References
\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\vspace*{-0.2in}
\noindent
-->
