---
author: 
  - name: Rolf Simoes, Gilberto Camara
    affiliation: INPE, Brazil
  - name: Victor Maus
    affiliation: IIASA
  - name: Alexandre Iwata
    affiliation: IPEA, Brazil
title:
  formatted: '\pkg{SITS}: An R Package for Data Access, Visualisation, Filtering,
    Clustering, Event Detection and Classification of Satellite Image Time Series'
  plain: 'SITS: Satellite Image Time Series package'
  short: SITS package
abstract: Using time series derived from big Earth Observation data sets is one of the leading research trends in Land Use Science and Remote Sensing. One of the more promising uses of satellite time series is its application for classification of land use and land cover, since our growing demand for natural resources has caused major environmental impacts. Given this motivation, this package provides a set of tools for data access, filtering, clustering and classification of satellite image time series.
keywords:
  formatted: [satellite image time series, big Earth Observation data]
preamble: >
  \usepackage{microtype} 
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: e-sensing.bib
---

# Introduction 
```{r, child='sec01_introd.Rmd'}
```



```{r include = FALSE}
library(sits)
```

\section{Using the WTSS service} 

Get information about the WTSS (web time series service)

To get a remote sensing time series, one first organises a large set of EO data as a 3D array. From each pixel location in the array, one can extract a time series of one or more variables for a temporal interval. The WTSS service is independent of the actual data architecture used for 3D array store. It can work with solutions such as flat files, MapReduce distributed datasets, array databases or object-relational databases. We have implemented the service using both a set of flat files and the SciDB array database management system \citep{Stonebraker2013}, with the same external interface. 
```{r}
URL <- "http://www.dpi.inpe.br/tws/wtss"
wtss_inpe <- sits_infoWTSS(URL)
```


```{r}
# get information about a specific coverage
sits_coverageWTSS(URL,"mod13q1_512")
# choose a coverage
coverage <- "mod13q1_512"
# recover all bands
bands <- c("ndvi", "evi", "nir")
# a point in the transition forest pasture in Northern MT
long <- -55.57320
lat <- -11.50566
# obtain a time series from the WTSS server for this point
series.tb <- sits_getdata(longitude = long, latitude = lat, URL = URL, coverage = "mod13q1_512", bands = bands)
# plot the series
sits_plot (series.tb)
```

