---
author: 
  - name: Rolf Simoes, Gilberto Camara
    affiliation: INPE, Brazil
  - name: Victor Maus
    affiliation: IIASA
  - name: Alexandre Iwata
    affiliation: IPEA, Brazil
title:
  formatted: '\pkg{SITS}: An R Package for Data Access, Visualisation, Filtering,
    Clustering, Event Detection and Classification of Satellite Image Time Series'
  plain: 'SITS: Satellite Image Time Series package'
  short: SITS package
abstract: Using time series derived from big Earth Observation data sets is one of the leading research trends in Land Use Science and Remote Sensing. One of the more promising uses of satellite time series is its application for classification of land use and land cover, since our growing demand for natural resources has caused major environmental impacts. Given this motivation, this package provides a set of tools for data access, filtering, clustering and classification of satellite image time series.
keywords:
  formatted: [satellite image time series, big Earth Observation data]
preamble: >
  \usepackage{microtype} 
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: e-sensing.bib
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{SITS: Satellite Image Time Series package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include = FALSE}
library(sits)
```


# Introduction 
```{r, child='sec01_introd.Rmd'}
```

# Data Handling in SITS
```{r, child='sec02_data.Rmd'}
```

# Using the Web Time Series Service WTSS
```{r, child='sec03_wtss.Rmd'}
```

\section*{Acknowledgments}
This work is supported by the São Paulo State Foundation (FAPESP) under grant 2014/08398-6 ("e-sensing: Big Earth observation data analytics for land use and land cover change information"), and by the Germany International Klimate Initiative under the RESTORE+ grant17_III_084_Global_A_RESTORE+ (“RESTORE+: Addressing Landscape Restoration on Degraded Land in Indonesia and Brazil”).  Gilberto Camara's work is also supported by CNPq (grant 312151\/2014-4).


